Using Visual Analytics to Explain\\[1ex] Black-Box Machine Learning

### 2018-03-20 tex/current.tex
some text

\chapter{Outlook}
\label{chap:current}
\todo{end???}
The model diagnostic workflow proposed in the previous chapter has so far only been tested with binary input features.
This poses the question on how to expand to general numeric features.
While it is possible to adapt the used explanation algorithm to accept arbitrary features the resulting explanations are less interpretable.

>>>>

\chapter{Outlook}
\label{chap:current}

\todo{see motivation for building context}
\todo{discussion? -- also overfitting}
Interestingly, because humans are used to having contextual knowledge about real world problems machine learning is trying to solve, it often seems surprising when and why a model failed, since the correct prediction ``is so obvious".
This also means, that repeatedly seeing that a model performs correctly, especially when providing a reasoning behind the decision in the form of \emph{explanations} might lead to overconfidence of the human in the capabilities of the model.
\todo{explanations considered harmful?}

\todo{end???}
The model diagnostic workflow proposed in the previous chapter has so far only been tested with binary input features.
This poses the question on how to expand to general numeric features.
While it is possible to adapt the used explanation algorithm to accept arbitrary features the resulting explanations are less interpretable.

### 2018-03-20 tex/introduction/machine_learning.tex
some text

One of the main advantages of logistic regression is its interpretability.
The weights $w_i$ can be directly interpreted as how influential a particular feature is.
That is, if the magnitude of a weight is large, a small change in the value $x_i$ of the feature has a big impact on the overall sum.
Additionally, the sign of a weight indicates whether a particular feature is positively or negatively correlated with the outcome.
For example, a simple (not necessarily correct) predictor for Type 2 Diabetes\footnote{A disease, affecting the body's ability to regulate blood sugar, which is mostly driven by lifestyle choices and associated with obesity and hypertension. The most common age of onset is between ${\sim}45$ and ${\sim}65$.} could look like:
\[
l_\text{diabetes}(x) = -1.9 + 0.01 x_\text{age} + 0.02 x_\text{mass} + (-0.3) x_\text{height}
\]
with $x_\text{age}$ in years, $x_\text{mass}$ in kg, and $x_\text{height}$ in m.
From here, we can see that the model assumes that a high mass leads to higher Diabetes risk, old age also leads to a higher risk but less so than high mass, and large body height leads to a smaller risk.
This gives a very intuitive understanding of how the model is reaching its conclusion.

@@

Decision Trees are usually considered very interpretable, as their behaviour is intuitively understandable.
However, this only holds true for trees with a relatively low number of nodes.
Trees with hundreds or thousands of nodes \todo{expand}

\subsubsection{Ensemble Methods}
\todo{random forest / ensemble methods}

@@

\subsection{Failure Modes}
\todo{failures}
- overfitting
- underfitting
- data
- model
- unexpected and different from how humans would fail

>>>>

One of the main advantages of logistic regression is its interpretability.
The weights $w_i$ can be directly interpreted as how influential a particular feature is.
That is, if the magnitude of a weight is large, a small change in the value $x_i$ of the feature has a big impact on the overall sum.
Additionally, the sign of a weight indicates whether a particular feature is positively or negatively correlated with the outcome.
For example, a simple (not necessarily correct) predictor for Type 2 Diabetes\footnote{A disease, affecting the body's ability to regulate blood sugar, which is mostly driven by lifestyle choices and associated with obesity and hypertension. The most common age of onset is between ${\sim}45$ and ${\sim}65$.} could look like:
\[
l_\text{diabetes}(x) = -1.9 + 0.01 x_\text{age} + 0.02 x_\text{mass} + (-0.3) x_\text{height}
\]
with $x_\text{age}$ in years, $x_\text{mass}$ in \si{kg}, and $x_\text{height}$ in \si{m}.
From here, we can see that the model assumes that a high mass leads to higher Diabetes risk, old age also leads to a higher risk but less so than high mass, and large body height leads to a smaller risk.
This gives a very intuitive understanding of how the model is reaching its conclusion.

@@

Decision Trees are usually considered very interpretable, as their behaviour is intuitively understandable.
However, this only holds true for trees with a relatively low number of nodes.
Trees with hundreds or thousands of nodes are hard to interpret as their complexity makes it hard to gain a mental model.

Decision Trees excel in cases where certain ranges of features have a clear impact on the outcome.
For example, in our running Diabetes use case, the risk of becoming diabetic is significantly higher for ages between ${\sim}45$ and ${\sim}65$.
A Decision Tree can then differentiate by age and infer the prediction differently for values inside or outside of that range.

However, Decision Trees cannot model relationships between features very well and have to approximate them.
For example, increasing the height of a person also increases that person's weight.
This means that whether a person is overweight, not only depends on the weight of the person, but also how tall that person is.
A person weighing $80\si{kg}$ can be considered overweight with a height of $1.7\si{m}$ but would not be with a height of $1.8\si{m}$.
Since a Decision Tree cannot model this relationship ($x_\text{mass} / x_\text{height}^2 \geq 25$) directly it has to approximate it by breaking the values of $x_\text{mass}$ and $x_\text{height}$ into intervals and checking them separately.
\todo{over-fitting}

\subsubsection{Ensemble Methods}
A powerful method of utilizing several weak models are Ensemble Models.
Instead of taking the prediction of one model, Ensembles combine the prediction of many models in order to get a definite result.
This is typically done by majority voting, \ie, the most common prediction among the individual models is used as final prediction.
\todo{cite papers}

The main advantage of Ensemble Models is that it can improve the performance of otherwise worse performing models.
Given a set of independent / uncorrelated models the error reduction of the Ensemble behaves linearly with number of individual models:
\todo{formula and citation -- chapter regularization}

One popular Ensemble Model is the Random Forest, which combines several Decision Trees trained on random sub-samples of the training data.
\todo{avoid over-fitting; better than Decision Tree}
\todo{link back to running example}

@@

\subsection{Failure Modes}
Machine learning models not always perform perfectly.
Most \todo{}
\todo{failures}
- overfitting
- underfitting
- data
- model
- unexpected and different from how humans would fail

### 2018-03-20 tex/introduction/motivation.tex
some text

\section{Motivation}
\todo{Motivation}
Modern machine learning models

* 3 failure modes: hidden bias, missing context, incomplete generalization
* failure modes different than for humans / surprising to humans
* 3 examples: pneumonia, radio, toaster
% \cite{Caruana:2015:IMH:2783258.2788613}

% \cite{Bird:2002:ERI:1251972.1252349}
% https://people.duke.edu/~ng46/topics/evolved-radio.pdf

% \cite{2017arXiv171209665B}
% https://arxiv.org/pdf/1712.09665.pdf

* reasons for black box
* gradient to explanation

>>>>

\section{Motivation}
\todo{Motivation}
Modern machine learning algorithms have reached a degree of potential that lets the resulting models model almost any dataset.
But even though there are statistical measures to guarantee that the model is learning correctly without over fitting on the training data, the model with the highest accuracy is not always the most usable model or even semantically correct.
The following three examples aim to showcase this:

\par \noindent (1)
\todo{pneumonia example}
% \cite{Caruana:2015:IMH:2783258.2788613}

\par \noindent (2)
\todo{radio example}
% \cite{Bird:2002:ERI:1251972.1252349}
% https://people.duke.edu/~ng46/topics/evolved-radio.pdf

\par \noindent (3)
\todo{toaster example}
% \cite{2017arXiv171209665B}
% https://arxiv.org/pdf/1712.09665.pdf

Those three examples all exhibit limitations of machine learning models where increasing the statistical accuracy of the model does not result in a semantically more correct model.
This is due to inherent biases of the data.

In the pneumonia example (1) patients with exceptionally high mortality risk were removed from the data leading to the model being unable to detect the severity of those cases.
In the radio example (2) the model utilized contextual data from the environment that is present but \emph{should} not be used in the model.
And in the toaster example (3) the model expected that all input data is coming from physical, real, objects.
Those biases cannot be detected from within the model or the data itself.
Furthermore, the process of detecting those errors, or even optimizing for semantic accuracy cannot be automated or be formulated as procedure, since there is no objective or measurable optimization criteria for it.
This is in contrast to statistical accuracy where a higher number always indicates a better model.
As a result, human judgment and contextual knowledge is absolutely necessary.

In order for a technique to be effectively useful for a human it needs to have few restrictions \todo{}
* reasons for black box
* gradient to explanation
